Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-06-22 ltx/PLoS/LaTeX_Manuscript.tex
added discussion to new template

\subsection*{Subsection 1}

\subsection*{Subsection 2}

\section*{Discussion}

@@

\begin{table}[!ht]
\caption{
\bf{The Unscented RTS Smoother}}
\begin{tabular}{|c|}
\multicolumn{1}{|p{16cm}|}{ 1. Forward initialisation} \\
$\hat{\mathbf x}_0, \mathbf P_0$ \\
\multicolumn{1}{|p{16cm}|}{   Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $, calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}} \\
$\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) \quad i=0, \dots, 2L$ \\
\multicolumn{1}{|p{16cm}|}{   Calculate the predicted state and the predicted covariance matrix} \\
$\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} $ \\
$\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e$ \\ 
\end{tabular}
\begin{flushleft}Table caption
\end{flushleft}
\label{tab:label}
\end{table}

\begin{algorithm}
	\begin{small}
	\caption{The Unscented RTS Smoother}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
		 \hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
		calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}
% 		\begin{small}
		\begin{equation*}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) \quad i=0, \dots, 2L
		\end{equation*}
% 		\end{small}
		calculate the predicted state and the predicted covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman Filter update equations
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\epsilon})^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation*}
			\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation*}
		\State Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propagate them through equation \ref{eq:QmatrixForSigmapoints}
		\begin{equation*}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b) \quad i=0, \dots, 2L
		\end{equation*}
		 compute the predicted state, the predicted covariance matrix and the cross-covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{b-})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{b-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{b-})^\top 
		\end{equation*}
		 Compute the smoother gain, the smoothed state and the smoothed covariance matrix
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{small}
\end{algorithm}

>>>>

\section*{Discussion}\label{DiscussionSection}
This paper has presented a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. As seen in the Section~\ref{sec:state_and_param_results} the results indicate good estimation performance, where the mean parameter estimates are in good agreement with the actual values. The estimates of the connectivity kernel parameters show relatively broad distributions, which is expected considering the system is driven by an unknown input, modelled as a random disturbance, and the observations are corrupted by noise. To overcome the effect of noise, an evoked-response paradigm should be considered when applying the estimation algorithm to electrophysiological recordings. Following such a paradigm, data can be epoched into a sufficient number of trials (realisations), where each trial is a response to a stimulus, such that a distribution of the parameter estimates can be formed.

Previous work from Schiff and Sauer~\cite{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter (UKF) for state estimation for discrete (in space and time) nonlinear spatio-temporal systems. Within their framework, each discrete spatial location forms an element of the state vector to be estimated with the UKF. A limitation of this framework lies in an increase in dimensionality with increases in the spatial resolution and in the physical size of the neural field. Increases in dimensionality lead to increases in uncertainty of the estimated state and problems associated with computational complexity. This paper builds on this previous study, where the estimator is constructed to approximate a continuous field and the dimensionality of the state vector is independent of the resolution of a spatial discretisation and is governed by the inherent spatial properties of the field. The frequency analysis methods presented in Section~\ref{SpectralAnalysisSection} provide a rigorous framework for the placement and the number of the basis functions required for the decomposition. This enables estimation of a larger field, with a lower dimensional state vector. The basis function decomposition also enabled the coupling to be described by an inner product rather than a spatial convolution. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

The frequency analysis and model selection tools presented in Section~\ref{SpectralAnalysisSection} specify the sensor arrangement that is required for capturing the dominant cortical dynamics. Importantly, these new tools can also be used to design electrophysiological recording systems to avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}.

To the authors' best knowledge, this is the first paper derive a continuum neural field model-based estimation framework for intracortical connectivity from electrophysiological data. The importance of the estimation framework can be illustrated by considering that the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion of water in grey matter~\cite{Assaf2008}.

The closest competitor to our `grey-box' integro-difference equation (IDE) based approach is perhaps the `black-box' multivariate auto-regressive (MVAR) model. Examples of MVAR models for estimating functional connectivity include Granger causality~\cite{Hesse2003}, the direct transfer function~\cite{Kaminski1991} and partial directed coherence~\cite{Sameshima1999}. A distinct advantage of the IDE-based technique proposed in this paper, is that the estimates of the connectivity structure are made independently of the number of sensors (assuming the field is sufficiently sampled). In addition, the number of parameters required to model the system is considerably less than a typical MVAR model. Furthermore, by using the neural field equations as a parametric form of the model, the parameters have physical meaning and therefore, can be clinically relevant. An advantage of the MVAR-based methods is that they allow for anisotropic connectivity, which is present in the longer range cortico-cortical fibres. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of a multi-scale connectivity structure in generating the characteristic rhythms of the brain. A possible method of creating more precise models with this multi-scale structure maybe to estimate anisotropic cortico-cortical connectivity using a MVAR approach and intracortical connectivity using the methods proposed in this paper.

Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

Parameter estimation may also have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from either abnormal excitatory or inhibitory mechanisms~\cite{Marten2009}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies.

The development of an estimation framework also has theoretical implications. The the ability to estimate parameters will allow testing of hypotheses that have been generated in theoretical studies. This has the potential to validate neural field models, and in doing so, provide new insights that may lead to a greater understanding of macro-scale neurodynamics. For example, using the proposed model-based framework we can reconstruct a neural field from data and estimate parameters and check if they correspond to theoretically derived parameters where certain solutions can exist. In addition, the parameters and the state-space can be explored during cognitive tasks and states of arousal such as during sleep and anaesthesia.

In this paper, an assumption was made that the chosen kernel basis functions can exactly represent the actual connectivity kernel. In future work, we plan to relax this assumption. We hope to achieve this by using the actual kernel that cannot exactly be presented by the chosen finite number of basis functions. In addition, we have assumed the neural field is isotropic and homogeneous. It is generally accepted that the intracortical connectivity structure can be accurately modelled as being isotropic. However, the assumption of homogeneity is quite strong. Future work should be focussed on developing estimation algorithms where assumption of homogeneity can be relaxed. Future work should also be targeted to include a more realistic synaptic response kernel, with finite rise and decay times and finite, and distant dependant propagation times into the parametric neural field equations.

% This paper presents a novel framework for state and parameter estimation of spatio-temporal neural fields to create patient specific models. A basis function decomposition reduces the infinite dimensional neural field (continuous space) to a finite dimensional state-space model. The frequency analysis methods presented in this paper provide a systematic procedure for specifying the sensor and basis function arrangement required to describe the significant neuro-dynamics by the reduced neural field model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch-Tung-Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. We have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic dynamics using the proposed methodology.

@@

\begin{table}[!ht]
\caption{
\bf{The Unscented RTS Smoother}}
\begin{tabular}{|c|}
\multicolumn{1}{|p{16cm}|}{ 1. Forward initialisation} \\
$\hat{\mathbf x}_0, \mathbf P_0$ \\
\multicolumn{1}{|p{16cm}|}{   Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $, calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}} \\
	$\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) \quad i=0, \dots, 2L$\\
\multicolumn{1}{|p{16cm}|}{   Calculate the predicted state and the predicted covariance matrix} \\
$\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} $ \\
$\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e$ \\ 
\multicolumn{1}{|p{16cm}|}{Compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman filter update equations} \\
$\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\epsilon})^{-1}$
\end{tabular}
\begin{flushleft}Table caption
\end{flushleft}
\label{tab:label}
\end{table}

\begin{algorithm}
	\begin{small}
	\caption{The Unscented RTS Smoother}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
		 \hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
		calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}
% 		\begin{small}
		\begin{equation*}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) \quad i=0, \dots, 2L
		\end{equation*}
% 		\end{small}
		calculate the predicted state and the predicted covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman filter update equations
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\epsilon})^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation*}
			\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation*}
		\State Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propagate them through equation \ref{eq:QmatrixForSigmapoints}
		\begin{equation*}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b) \quad i=0, \dots, 2L
		\end{equation*}
		 compute the predicted state, the predicted covariance matrix and the cross-covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{b-})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{b-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{b-})^\top 
		\end{equation*}
		 Compute the smoother gain, the smoothed state and the smoothed covariance matrix
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{small}
\end{algorithm}
